Matching the Time of Emergence of Transboundary Fish Stocks to Lead Time for

### 2020-09-04 Scripts/Manuscript.Rmd
Sent first draft to W. up to results

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species are changing their distribution [@ Poloczanska:2016kk] and will continue to do so in the search of the apropiate environmental conditions needed for their subsistence [@Cheung:2010dt. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@SDG14; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @CheungClimateReadyFish]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg;@Pinsky:2018cb;].

The concept of shared stocks gained attention in the early 1980's when the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl]. The implementation of arbitrary jurisdictions across the global ocean created a partition in the distribution of several stocks, thus, creating what we known today as shared stocks. As defined by the FAO, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover broad oceanic regions [@FAO; @Munro]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence suggest that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@EuropeShift]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song]. In the South-eastern Pacific, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted south (towards Chilean waters) adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. We argue that two concepts will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries.

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. Thus, if a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S].This can be seen as the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@PalaciosAbrantes:E&S]. Many shared fisheries operate this way [@EuropeShift; @GoMA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@PalaciosAbrantes:E&S; @Sumaila:E&S]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive biomass further away from the historic range limits of a species [@Alexa?]. Assessing the ToE of shared stocks is as important as the change itself as it could reduce potential international conflicts [@Spijkers:2017ij] and prepare for emerging fisheries [@Pinsky:2018cb;  @Pinsky:2014by].

@@

The Food and Agricultural Organization (FAO) of the UN define transboundary stocks as those that freely swim across the EEZ of neighboring nations [@REFFAO]. Accordingly, we included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs [@Palacios-Abrantes:xxxx]. We consider each species as a single stock unit within an EEZ due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members [@Rodggers:2015jq].

## Changes in Stocks Share Ratio

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@Palacios-Abrantes:E&S]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@Palacios-Abrantes:xxxx]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. We then estimate the percentage change in the SSR of each future time period relative to the historic time period. To address the natural spatial variability of the stock we only considered a SSR change when the future SSR was higher or lower than the average historic SSR $\pm$ 1 s.d. The analysis was replicated for each of the ten ensemble members capturing variation by averaging results among ensemble members (mean $\pm$ standard deviation).

Knowing the time (e.g. year) in which a shared stock is going to diverge from a historical variability can be key for preparing to shifting stocks. We employed the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

@@

$$TI =(\frac{D_{A,t}}{sd(D_{A,h})} - \frac{D_{B,t}}{sd(D_{B,h)}})^2$$

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*h*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, to determine the time of emergence we estimated a signal to noise ratio (SNR) and set an arbitrary threshold as follows [@Rodggers:2015jq]:

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated following the SNR method.

@@

We tested data for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated changes in the SSR of transboundary species shared by neighboring EEZs by early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). While most EEZs will experience an average change lower than 25%, a few EEZs are expected to have larger changes in SSR. Such change is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. The number of stocks per EEZ that will see a change in SSR by early $21^{st}$ century is relatively low (median of 13 spp. per EEZ), however, some regions like West Africa and the Mediterranean will see changes in over 50 species (e.g. Senegal, 94). These regions, where EEZs are relatively small and have multiple borders will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^st$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to early century, but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

![Figure 1. Average gain of transboundary stock share ratio (lines) and number of shifting stocks (polygon) by 2030 (2021-2040) relative to 1951-2005. Arrows go from the losing EEZ (point) to the winner (arrowhead). Panel B zooms in to specific areas.](../Figures/Fig1.png)

@@

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock's distribution overpasses the historical variation. Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). The cumulative sum of emerging stocks will have a steeper increase until 2025 where the increment will be slightly less intense (Fig. 2). Similarly, for the number of EEZs, the increment curve seem to flatten by 2020. By 2081, all of the EEZs analyzed here will have at least one emerging stock while 2097 will be the year where all 633 stocks have emerged in at least one neighboring EEZ (Fig. 2).

@@

![Figure 2 Cumulative increase in time of emergence by EEZ and stock](../Figures/Fig2.png)

Most of the world EEZ’s will see an average ToE across transboundary stocks by 2030 with a few EEZs having a relative late average ToE (e.g. Honduras average ToE is 2071) and 54 EEZs having already emerged (e.g. average ToE <= 2020). A higher number of emerging stocks will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50%. However, in some cases emerging stocks represent 85% of total fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, no stock will emerge between 2006 and 2100.

@@

![Figure 3. Average time of emergence (ToE) per Exclusive Economic Zone (EEZ polygon) and participation of emerging stocks in total EEZ fishing revenue (land polygon). Darker tones represent low participation and late ToE; lighter colors represent large participation and earlier ToE. Density plot on right shows the number of emerging stocks across EEZ centroids (10^o running mean).](../Figures/Fig3.png)

Most fisheries within the EEZ of the main transboundary fishing countries in terms of revenue (e.g. the top $75^{th}$ percentile) will emerge after 2030 (n = 28). The remaining twelve countries will have an average ToE before 2030, this includes the United States who is the $4^{th}$ place in total revenue (Fig. 4). These 47 countries are also among those with the largest number of emerging taxa. Regionally speaking, most of African, Asian and European EEZs will see a latter ToE, while the Americas and Oceania will see an early ToE (Fig. 4). From the top ten countries with more revenue from transboundary species, nine of them share a border. Most of these countries (e.g. Spain and China) will also have a large number of emerging stocks, while Peru and Norway will have a small number of emerging taxa but representing high revenue fisheries.

@@

![Figure 4. Average time of emergence (ToE) per fishing entity. Fishing revenue in logarithmic scale Showing names for the top 75th percentile). Countries marked with an asterisk have at least one border.](../Figures/Fig4.png)

We compared species ToE within EEZs by geographic region and by species habitat preference. The mean ToE of transboundary species varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, *DF* = 93, *p* < 0.001) (Fig. 5). Post-hoc tests revealed that EEZs in Easter Asia which have the latest average ToE across all habitat types (2041), and Latin America and the Caribbean and Polynesia who have the earliest average ToE across all habitats (2021 and 2018, respectively), significantly differ from (almost) all other regions. Other regions also significantly differ from each other but in a less frequent manner (*See Table Sup X for full statistics*). When looking specifically at habitat association, we found significant differences on ToE of different habitat types (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (*See Table Sup XX for full statistics*).

>>>>

Over the last century human activities have fundamentally changed the physicochemical composition of the ocean, warming the waters, increasing the pH and reducing the oxygen concentration [@IPCC:2019to]. As a consequence, many marine species have changed their distribution [@Poloczanska:2016kk] and will continue to do so in the search of the appropriate environmental conditions needed for their subsistence [@Cheung:2010dt]. Climate change is expected to impact fisheries production at different latitudes compromising our capacity to reach international sustainability goals such as the SDG 14 - life below water [@SDG14; @Cheung:2010dt; @Barange:2014kr]. Such impacts can be offset with the proper reform of current fisheries management [@Gaines:2018sg; @CheungClimateReadyFish]. However, such reforms will require comprehensive plans and institutions to manage species that cross international borders, also known as shared stocks [@Gaines:2018sg;@Pinsky:2018cb;].

In the early 1980's the United Nations officialized the concept of Exclusive Economic Zone (EEZs) though the Law of the Sea (UCLOS) [@UnitedNations:1986tl] partitioning the distribution of several species, thus, creating what we known today as shared stocks. As defined by the Food and Agriculture Organization, shared stocks can be classified as transboundary, those that cross neighboring EEZs; straddling, that, in addition to neighboring EEZs, also visit the adjacent high seas; and highly migratory, mainly tunas and bill-fishes that cover vast oceanic regions [@FAO; @Munro]. While counties are responsible for the management of stocks within their EEZs, UNCLOS incentives actions to cooperate when stocks are shared [Article 63, @UnitedNations:1986tl]. Today, an estimate of 633 transboundary fished stocks are found in different continents and across multiple habitats averaging 48.5 million tones and representing USD 78 billion in fishing revenue between 2005 and 2010 [@FishorVisa].

The management, and thus sustainability, of fisheries that target shared stocks is jeopardized as marine species shift their known distribution to cope with a changing world [@Pinsky:2018cb]. There is evidence that such shifts have already complicated shared marine resources around the globe. In the North-eastern Atlantic, evidence shows that shared stocks managed within the European Union divisions have been shifting since the 1970's compromising the region's sustainability [@EuropeShift]. In the North-eastern Pacific, sockeye salmon (*Oncorhynchus nerka*) shifted in the 1980's towards Alaska, creating a ten years conflict between Canada and the United States [@Song]. Further south, the southern stock of Peruvean anchoveta (*Engraulis ringens*) shared by Chile and Peru has gradually shifted towards Chilean waters adding to the over-exploitation of the resource [@Cashion:2018cg]. Global shifts in the distribution of tunas [@ErauskinExtramiana:2019fw] have brought additional management challenges in both the Western Indian ocean and South-western Pacific [@Miller:2007kx]. Moreover, as species continue shifting with climate change, current policies might become obsolete, and thus, require new management strategies that create equitable systems across all dependent fishing nations [@Bell:2013ga; @Pinsky:2018cb]. The objective of the current paper is to explore two concepts that will be essential in addressing the challenges that climate change will bring to the management of transboundary fisheries [@Link:2010ei].

The first concept alludes to the game theory concept of thread point, known as the minimum payoff a player is willing to receive in order to cooperate with others. If a player's future payoff drops below the thread point, such player will likely break off any collaboration agreement and proceed to act solo [@Sumaila:E&S]. A thread point can be translated to the proportion of a species distribution that each neighboring nation has (or the quota allocation that each one gets) at the time of signing off an agreement [@PalaciosAbrantes:E&S]. Many shared fisheries operate this way [@EuropeShift; @GoMA; @NOAA]. If not addressed, future changes in the thread point (hereafter called Stock Share Ratio -SSR) of jointly managed fisheries can have adverse consequences for shared fisheries [@PalaciosAbrantes:E&S; @Sumaila:E&S]. The second concept refers to the Time of Emergence (ToE), defined as the moment when a future trend of a particular variable exceeds the historical variability [@IPCC:2019to]. Such concept has been used to look at the emergence of diverse oceanic variables [@Roggers:2015jq; @Keller:2014jv; @Frolicher:2016dg] and ecological consequences of marine heatwaves [@Cheung:2020ja]. Similarly, climate change is expected to drive species further away from their historic range limits [@Alexa?]. Assessing the ToE of shared stocks is as important as the change itself, as it could reduce potential international conflicts [@Spijkers:2017ij] and aid the strengthen of marine governance [@Pinsky:2018cb;  @Pinsky:2014by; @Link:2010ei].

@@

We included in this analysis 633 marine exploited transboundary species representing 80% of the catch taken from the world's EEZs between 2005 and 2014 [@Palacios-Abrantes:xxxx]. Each species in within an EEZ was considered a single stock unit due to a lack of more detailed spatial information [@Teh:2015gd; @Palacios-Abrantes:xxxx]. We defined the boundaries of the world EEZs (e.g., Mexico Pacific EEZ and Mexico Atlantic EEZ) using the *Sea Around Us* shapefile (updated 1 July 2015, available from http://www.seaaroundus.org) noting that it subdivides the EEZs of 198 coastal states into 280 regions (e.g., Mexico's EEZ is divided in Mexico Pacific and Mexico Atlantic) including islands territories, and determined the intersections between polygons using the R package *sf* (**Supplements Fig 1 (SuppM)**) [@PackagesfSimpleF:2018vp]. For each species and EEZ we used the Sea Around Us data to estimate the catch and fishing revenue (USD 2019 from fishing activities within the global EEZs reported as the average from the last decade (2005-2014) of available spatial data [@Zeller2016; @Tai2017; @Sumaila:2015uc]. Moreover, we classified catch trends as increasing, constant and decreasing, following Palacios-Abrantes *et al.* (*hopefully in press*). Finally, each EEZ was attributed to a geopolitical region [@UN] and biome [@Reygondeau:2019uh], and species were attributed a habitat preference [@FishBase].

@@

We coupled the DBEM with the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project species distribution from 1951 to 2100 under climate change [@Dunne:2012bo; @Dunne:2013db]. We run the DBEM simulating biomass and MCP from 1950 to 2100 under a “worst-case” climate change scenario according to the Representative Concentration Pathway 8.5 (RCP 8.5) used by the Intergovernmental Panel on Climate Change (IPCC) [@IPCC:2014bz; Riahi:2011dk]. According to the GFDL-ESM2M, under the RCP 8.5 scenario atmospheric temperature will increase by $3.2^{\circ}C$ by 2081-2100, relative to preindustrial levels. Because the main approach of this paper relies on understanding the spatial and temporal variation of a species distribution, we have to understand both the historical as well as the future climate-driven distribution variability, in order to infer the differences. We do this by following an ensemble member approach where each member constitutes a different initial Earth system condition [@Rodggers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, in each of the ten GFDL-ESM2M-ensembles the Earth system initial condition is infinitesimally different resulting in a unique atmosphere and ocean state at each point in time. This results in a different state of internal variability for each ensemble representing a plausible future under RCP 8.5 [@Rodggers:2015jq]. Following this approach allows us to estimate a secular trend for the natural variability (e.g. background noise). At the same time, by having multiple initial-condition ensemble simulations of the GFDL-ESM2M we were able to remove the natural variability of a stock's distribution by averaging the secular trend over all ensemble members (See Estimating the Time of Emergence). 
## Changes in Stocks Share Ratio

We adopted the concept of threat point as the minimum proportion of the stocks’ joint distribution between neighboring EEZs that each country will accept if they were to cooperate in the management of the stock [@Palacios-Abrantes:E&S]. For each stock and time step (1951-2100) we first aggregate the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells where the species was present within neighboring EEZs, and then calculated the proportion that each EEZ had [@FishForVisa]. We then average the results in three time periods in order to reduce climate variability within the projections. The first time period ($t_h$) comprises (1951 - 2005) and is assumed as the historic distribution of each species, thus representing each neighboring nation's threat point. This time frame matches the historic climate data that the GFDL uses to project climate change [@Dunne:2012bo; @Dunne:2013db]. We then selected two future periods; early $21^{st}$ century as the average of 2021 to 2040 ($t_e$), and mid $21^{st}$ century as the average of 2041 to 2060 ($t_m$). We choose these time periods to match the challenges of achieving fisheries-related UN-SDGs [@UnitedNations2018] such as SDG 14.4 (end overfishing) and SDG 1.2 (poverty reduction), aimed to be achieved by 2020 and 2030, respectively [@Singh:2017ds]. The analysis was replicated for each of the ten ensemble members capturing the variation by averaging results among ensemble members ($\mu \pm \sigma$). We then estimate the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_{f}$) relative to the historic time period ($SSR_h$) as follows,

$$ \Delta{SSR_{e,m}} = \frac{(SSR_{f}-SSR_{th})}{SSR_{th}}* 100 $$
To address the natural spatial variability of the stock we only considered $\Delta{SSR_{e,m}}$ to be true when $SSR_f \geq (SSR_f + \sigma)$ or $SSR_f \leq (SSR_f - \sigma)$.

## Estimating the Time of Emergence

Knowing the time (e.g. year) in which a shared stock is going to diverge from a historical variability is key for preparing to shifting stocks [@Link:2010ei]. We employed the DBEM and the ten GFDL ensemble members described above, and developed an index of transboundary range shift, fisheries and climate change (*TI*). The index is based on the variation of a species distribution centroid in relation to the centroid of the neighboring EEZs it occupy. Each species centroid was determined by the average latitude and longitude of the grid cells containing the top 10% abundance percentile of each year derived from the DBEM. The EEZ centroid was estimated using the *st* package of *R* statistical software (**Supplements Fig 1 (SuppM)**). For each ensemble member, neighboring EEZ and species, we computed the distance between centroids assuming a perfect sphere and ignoring elevation differences using the *geosphere* package of *R* statistical software;

@@

$$TI =(\frac{D_{A,t}}{sd(D_{A,th})} - \frac{D_{B,t}}{sd(D_{B,th)}})^2$$

where $D_A$ and $D_B$ represent the distance between EEZ and distribution centroids of a species in EEZs A and B, respectively for each future time step from 2006 to 2100 (*t*); *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. The index was then smoothed to a 10 years average. Finally, to determine the time of emergence we estimated a signal to noise ratio (SNR) and set an arbitrary threshold as follows [@Rodggers:2015jq]:

@@

where *TREND* is the yearly mean *TI* ($\mu$) of all ten ensemble members and *NOISE* is the standard deviation of the TREND ($\sigma$) between 1951 and 2005. Note that in some cases not all ensemble members show the presence of a species in a certain EEZ throughout the study's time frame. We presented the results only for those cases where all 10 ensemble members had results for each year between 1951 and 2100 (n = 150). We determine a $\mu \pm \sigma$ (confidence interval of 68%) to show the first year when the "signal" of the transboundary index overpasses the historic "noise". Since this was an arbitrary decision, we have included a more conservative estimation ($\mu \pm 2*\sigma$ for a 95% confidence interval). The SNR for both bottom and surface sea temperature was also estimated for statistical analysis following the SNR method.

@@

We tested the results for normality (e.g. skewness, kurtosis) and performed a couple of non-parametric Kruskal–Wallis test by ranks [@Hollander:2013wz] to investigate geopolitical and ecological differences in the ToE of transboundary stocks. Moreover, we implemented a multiple linear regression analysis to a series to explain the ToE of each species in each EEZ. Specifically, we looked at habitat preferences, catch trends, geopolitical and biome location and ToE of both bottom and sea surface temperature (See *Databases and species selection*). We hypothesize that all of these factors would have significant influence in the ToE of transboundary stocks. All analyses were run using the statistical software *R version 3.5.2 (2018-12-20; Eggshell Igloo)* with the packages *data.table* [@Packagedatatable:2019uh], *janitor* [@PackagejanitorSim:2018ur], *rfishbase* [@PackagerfishbaseR:2019th], *sf* [@PackagesfSimpleF:2018vp], *sp* [@PackagespClasses:2019tn], *tidiverse* [@PackagetidyverseE:2017vq], *tidytext* [@PackagetidytextTe:2019vk], *gmt* [@gmt] and *zoo* [@PackagezooSInfr:2019tm], *readxl* [@REF], *ggrepel* [@REF], *zeallot* [@REF], *viridis* [@], *cowplot* [@], *moments* [@REF] and *pgirmess* [@REF]. All code is available at https://github.com/jepa/TransEmergence

@@

We estimated the change in SSR (larger than 1 s.d) of transboundary species shared by neighboring EEZs by the early (2020-2040) and the mid $21^{st}$ century (2040-2060), relative to present (1951-2005). Our results suggest that `r n_eez_per`% (n = `r n_eez`) of the world EEZs will see a change in SSR by 2030 (Fig. 1A). Most EEZs will average a change lower than 25% which is expected to slightly increase in intensity towards the  mid $21^{st}$ century (**Supplements Fig 1 (histogram)**). However, this result masks larger changes in SSR for specific species. For example, Guatemala's SSR of *Panulirus gracilis* shared with Mexico is projected to go from 12% to 26%. By early $21^{st}$ century, the percentage of transboundary stocks per EEZ that will see a change in SSR >< 1 s.d. of historical distribution will be relatively large (global mean of 69% of species in an EEZ). For 17 EEZs, 100% of their analyzed transboundary stocks will shift, while and only six EEZs will have less than 25% shifting stocks (Fig. 1A-Polygons). Regions where EEZs are relatively small and have multiple borders (e.g. The Caribbean) will see a complex dynamic of changes in SSR with no clear patterns (Fig. 1B). Towards the mid $21^{st}$ century, changes in SSR per EEZ will no be substantial (**Supplements Fig 1 (map)**) relative to the early $21^{st}$ century but are expected to increase in intensity (**Supplements Fig 1 (histogram)**).

@@

![Figure 1. Changes in Stock Share Ration by 2030 (2021-2040) relative to 1951-2005. Lines represent the average gain of transboundary stock share ratio with srrows going from the losing EEZ (point) to the winner (arrowhead). Polygons represent the percentage of initial stocks that will see a shift. Panel B zooms in to specific areas shown in grey boxes of A.](../Figures/Fig1.png)

@@

We estimated the ToE between neighboring EEZs for 633 transboundary stocks, that is, the first year where the future stock's distribution (2006-2100) overpasses the historical variation (1951-2005). Since 2006, 87% of the world EEZs have already seen at least one emerging stock and 60% of the transboundary species analyzed have emerged in at least a set of neighboring EEZ (e.g. *Engraulis ringens* shared by Chile and Peru emerged in 2010). Stocks will continue to emerge steadily towards 2097 when all 633 stocks will have had emerged in at least one neighboring EEZ (Fig. 2). The number of EEZs with emerging stocks will increase steeply until 2020 when it seems to flatten approaching 2081 when all of the EEZs analyzed here will have at least one emerging stock (Fig. 2).

@@

![Figure 2. Cumulative increase in time of emergence by Exclusive Economic Zone and transbounudary stock.](../Figures/Fig2.png)

Most of the world EEZ’s will see an average ToE across transboundary stocks before 2040 (Fig. 3). In a few EEZs stocks will have a relative late average ToE (e.g. Honduras average ToE is 2071) while 54 of the 280 EEZs (19%) have already seen their stocks emerged (e.g. average ToE across all stocks <= 2020). A higher number of emerging stocks will happen between latitudes 0$^o$ and 50$^o$N corresponding to regions of relatively small EEZs with multiple borders (e.g. The Caribbean and the Mediterranean) (Fig 2). In most cases, the current participation of the emerging stocks in a country's total fishing revenue is less than 50% (Fig. 2 - Polygons). However, in some cases emerging stocks represent 85% of total fishing revenue (e.g. Peru) or more (e.g. Marshall Islands, 92%). Only in very few cases like Brazil and Alaska's polar EEZ, no stock will emerge between 2006 and 2100.

@@

![Figure 3. Average time of emergence of transboundary stock. Land polygon represents  the current participation of emerging stocks in the country's fishing revenue. Exclusive Economic Zone polygon represents the average time of emergence across all stocks. Density plot on the right shows the latitudinal number of emerging stocks across EEZ centroids (10^o running mean).](../Figures/Fig3.png)

Twelve of the 47 countries responsible for 75% of the yearly revenue generated from transbondary stocks between 2005 and 2010 will see an average ToE before 2030 (Fig. 4). This includes the United States who occupies the $4^{th}$ place. These 47 countries are also among those with the largest number of emerging taxa. Regionally speaking, most of African, Asian and European EEZs will see a latter ToE, while the Americas and Oceania will see an early ToE (Fig. 4). From the top ten countries with more revenue from transboundary species, nine of them share a border. Most of these countries (e.g.China) will also have a large number of emerging stocks, while others like Peru and Norway will have a small number of emerging taxa but representing high revenue fisheries.

@@

![Figure 4. Average time of emergence (ToE) per fishing entity. Fishing revenue in logarithmic scale. Showing names for the top 75th percentile. Countries marked with an asterisk have at least one neighbouring border.](../Figures/Fig4.png)

We compared species ToE within EEZs by geographic region and by species habitat preference (Fig. 5). The median ToE of transboundary species varied significantly according to the geographic region of the neighboring EEZs (Kruskal-Wallis, $X^2$ = 287.23, *DF* = 93, *p* < 0.001). Post-hoc tests revealed regional patterns. EEZs in Easter Asia which have the latest average ToE across all habitat types (2041), and Latin America and the Caribbean and Polynesia who have the earliest average ToE across all habitats (2021 and 2018, respectively), significantly differ from (almost) all other regions. Other regions also significantly differ from each other but in a less frequent manner (*See Table Sup X for full statistics*). When looking specifically at habitat association, we found significant differences in ToE by habitat preference (Kruskal-Wallis, $X^2$ = 286.48, *DF* = 93, *p* < 0.001). Specifically, the ToE of species with preferences for pelagic-oceanic and reef habitats was significantly different than other habitat preferences (*See Table Sup XX for full statistics*).
